---
title: "Phenoarch platform - Cleaning procedure - Curve level - CARBayesST library"
author: "I.Sanchez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Phenoarch platform - Cleaning procedure - Curve level - CARBayesST library}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!---
Program  : CARBayesSTReport.Rmd
Author   : I.Sanchez (MISTEA)
Objective: a vignette for phisStatR library
           Use of spatio-temporal bayesian anova modelling to detect outlier curve on each genotype-scenario
           of an experiment, taking into account the lattice shape of the greenhouse
           require phisStatR library and fitCARBayesST() function
           The user has to adapt to his own analysis.
Creation : 21/06/2017
Update   : 18/10/2019
-->

***
# Objective 

Detection of outliers in time courses of an experiment in PhenoArch greenhouse. In this vignette, we use a toy data set of the *phisStatR* library (anonymized real data set). 

The example is conducted on plantHeight parameter, we modelise the plantHeight taking into account the temporality (thermal time) and the spatiality (lattice structure) with a bayesian spatio-temporal ANOVA model [3].

To detect outlier points, we retrieve the standardised residuals computed by the model and a point is an outlier if abs(res) > *threshold* (here threshold==4). the user will be able to change this threshold.

We produced the output of the model as well as some graphics.

# Import of data
The input dataset must contain some predefined columns, have a look to the structure of *mydata*:
```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  library(lubridate)
  library(dplyr)
  library(phisStatR)

  myReport<-substr(now(),1,10)

  mydata<-plant1
  str(mydata)
```

# CARBayesST library

A bayesian approach to model these data where the spatio-temporal structure is modelled via sets of autocorrelated random effets. Conditional autoregressive (CAR) priors and spatio-temporal extensions thereof are typically assigned to these random effects to capture the autocorrelation, which are special cases of a Gaussian Markov random Filed (GMRF) [3].

*ST.CARanova()* decomposes the spatio-temporal variation into 3 components:

* an overall spatial effect common to all time periods
* an overall temporal trend common to all spatial units
* a set of independent space-time interactions

we can add others factors (here the scenario (quali) ...)

The spatio-temporal auto-correlation is modelled by a common set of spatial random effect and a common set of temporal random effects and both are modelled by the CAR prior (Conditional AutoRegressive).

**Description of the results's table**:

* **Median**: posterior median estimation
* **2.5 and 97.5%**: 95% credible interval
* **Geweke.diag**: the convergence diagnostic implemented in the **coda** package. takes form of a Z-score, so the convergence is suggested by the statistic being **within** the range (-1.96; 1.96)

**Model fit criteria**

* DIC: deviance information criterion
* p.d: its corresponding estimated effective number of parameters
* WAIC: the Watanabe-Akaike information criterion
* p.w: its corresponding estimated number of effective parameters
* LMPL: the log marginal predictive likelihood

In model's comparison, the best fitting model is the one that minimises the DIC and WAIC but maximises the LMPL.

**As the bayesian model can take a while, in this example, we initialise burnin=10 and n.sample=110 - not enough in real analysis!**

```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  # We suppress observations with missing data in time variable (here thermalTime)
  mydata<-filter(mydata,!is.na(mydata$thermalTime))
  model<-fitCARBayesST(datain=mydata,xvar="thermalTime",trait="plantHeight",k=2,
          graphDist=TRUE,burnin=10,n.sample=110,
          formulaModel=as.formula(plantHeight~scenario+genotypeAlias),typeModel="anova",verbose=TRUE)

```

```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  # print the result of the bayesian modelling
  printCARBayesST(modelin=model[[1]])
```

```{r , echo = TRUE,message=FALSE, warning=FALSE}
  outtmp<-outlierCARBayesST(modelin=model[[1]],datain=model[[2]],threshold=4,trait="plantHeight")
```

## Outlier detection
The output report can be over-sized (more than 1Mb), for size of sub-directories in packages purposes, I choose to represent only the first genotypes...
```{r , echo = TRUE,message=FALSE, warning=FALSE,fig.height=10,fig.width=12}
  mygeno<-as.character(unique(model[[2]][,"genotypeAlias"]))
  mygeno<-mygeno[1:6]
  for (i in seq(1,length(mygeno),by=15)){ 
      myvec<-seq(i,i+14,1)
      plotCARBayesST(datain=model[[2]],outlierin=outtmp,myselect=myvec,trait="plantHeight",xvar="thermalTime")
  }
  
```

# Session info
```{r session,echo=FALSE,message=FALSE, warning=FALSE}
  sessionInfo()
```

# References
1. R Development Core Team (2015). R: A language and environment for statistical computing. R Foundation for
      Statistical Computing, Vienna, Austria. ISBN 3-900051-07-0, URL http://www.R-project.org.
2. Roger S. Bivand, Edzer Pebesma, Virgilio Gomez-Rubio, 2013. Applied spatial data analysis with R, Second edition. Springer, NY.
  http://www.asdar-book.org/
3. Duncan Lee, Alastair Rushworth and Gary Napier (2017). CARBayesST: Spatio-Temporal Generalised Linear Mixed Models for Areal Unit Data. R package version 2.5. https://CRAN.R-project.org/package=CARBayesST
 
