---
title: "Phenoarch platform - Cleaning procedure - Curve level using SpATS"
author: "I.Sanchez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Phenoarch platform - Cleaning procedure - Curve level using SpATS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!---
Program  : detectOutlierCurves.Rmd
Author   : I.Sanchez (MISTEA)
Objective: a vignette for phisStatR library
           require phisStatR library and () function
           The user has to adapt to his own analysis.
Creation : 04/05/2018
Update   : 18/10/2019
-->

***
# Objective 

```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  library(dplyr)
  library(tidyr)
  library(phisStatR)
  library(ggplot2)
```

This vignette deals with the detection of outlier plants in a lattice experiment using a spatial model using splines (**SpATS** library) [3]. Here the following steps of this procedure developed for Maize experiment but easily adaptable to others species:

* From a temporal dataset designed as __plant1__ with biomass (or biovolume) and plant height phenotypes. Extract predictions of these two phenotypes at a specific time point (for example just before the treatment) and create a dataset designed as __plant4__. Here, we will use the biomass and plant height predicted at 24 days et 20°C.
* From a temporal dataset of count of number of leaves, extract the phyllocron (slope of linear regression for each plant)

We have so a dataset with one row for each plant in the experiment containing the three phenotypes: __biomass24__, __PH24__ and __Phy__.

* Apply a SpATS model on each phenotype, check the diagnostic graphics and retrieve the deviance residuals calculated by the model
* On the residuals, we can detect outlier plant(s) with a combined physiological criterion applying the following rules:

* __raw procedure__: at a threshold=0.95 (can be modified)
    + __small plants__ are identified if $res_i < \mu_{res} - qnorm(threshold) \times sd_{res}$ for <span style="color:green"> biomass AND phyllocron </span>
    + __big plants__ are identified if $res_i > \mu_{res} + qnorm(threshold) \times sd_{res}$  for <span style="color:green">biomass AND plant height </span>

# Import of data
In this vignette, we use a toy data set of the *phisStatR* library (anonymized real data set). the first one *plant4* is extracted from *plant1* and is used to detect outlier at a specific time point. The detected outlier will be checked on the *plant1* dataset graphically.

```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  mydata<-plant4
  str(mydata)
```

# SpATS library
## Biomass
```{r spatmodel1,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  spat.biomass<-fitSpATS(datain=mydata,trait="Biomass24",genotypeId="genotypeAlias",rowId="Line",colId="Position",
                 typeModel="anova",genotype.as.random=FALSE,nseg=c(14,30),verbose)
```

```{r spatplot1,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  plot(spat.biomass)
```

## Plant height
```{r spatmodel2,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  spat.ph<-fitSpATS(datain=mydata,trait="PH24",genotypeId="genotypeAlias",rowId="Line",colId="Position",
                 typeModel="anova",genotype.as.random=FALSE,nseg=c(14,30),verbose)
```

```{r spatplot2,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  plot(spat.ph)
```

## Phyllocron
```{r spatmodel3,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  spat.phy<-fitSpATS(datain=mydata,trait="Phy",genotypeId="genotypeAlias",rowId="Line",colId="Position",
                 typeModel="anova",genotype.as.random=FALSE,nseg=c(14,30),verbose)
```

```{r spatplot3,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  plot(spat.phy)
```


# Data cleaning
The data cleaning procedure is to identify plants having too large or too small features, from deviance residuals obtained with SpATS mixed models for each parameter (biomass, plant height and phyllocron). 2 procedures are tested:

* __raw procedure__: at a threshold=0.95 or 0.90 or 0.98 (can be modified)
    + __small plants__ are identified if $res_i < \mu_{res} - qnorm(threshold) \times sd_{res}$ for <span style="color:green"> biomass AND phyllocron </span>
    + __big plants__ are identified if $res_i > \mu_{res} + qnorm(threshold) \times sd_{res}$  for <span style="color:green">biomass AND plant height </span>
* __quantile procedure__: Same procedure with quantiles : 
    + __small plants__ are identified if $res_{i} < Q1_{res} - 1.5 \times(Q3 - Q1)$ for <span style="color:green"> biomass AND phyllocron </span>
    + __big plants__ are identified if $res\_{i} > Q1_{res} - 1.5 \times(Q3 - Q1)$ for <span style="color:green"> biomass AND plant height </span>

```{r proc,echo=TRUE,message=FALSE, warning=FALSE}
  # concatenate raw data and residuals calculated with the SpATS model
  devResBio<-spat.biomass$residuals
  devResPH<-spat.ph$residuals
  devResPhy<-spat.phy$residuals
  myglobal<-cbind.data.frame(mydata,devResBio,devResPH,devResPhy)
  threshold<-0.95
  
  myglobal<-mutate(myglobal,mean.devBio=mean(devResBio,na.rm=TRUE),
                          sd.devBio=sd(devResBio,na.rm=TRUE),
                          mean.devPH=mean(devResPH,na.rm=TRUE),
                          sd.devPH=sd(devResPH,na.rm=TRUE),
                          mean.devPhy=mean(devResPhy,na.rm=TRUE),
                          sd.devPhy=sd(devResPhy,na.rm=TRUE),
                          lower.devBio=mean.devBio - sd.devBio*qnorm(threshold),
                          lower.devPH=mean.devPH - sd.devPH*qnorm(threshold),
                          lower.devPhy=mean.devPhy - sd.devPhy*qnorm(threshold),
                          upper.devBio=mean.devBio + sd.devBio*qnorm(threshold),
                          upper.devPH=mean.devPH + sd.devPH*qnorm(threshold),
                          upper.devPhy=mean.devPhy + sd.devPhy*qnorm(threshold),
                          # yes=1 == OK
                          # no =0 == problem
                          lower.critraw.spatsBio=ifelse(devResBio - lower.devBio > 0,yes=1,no=0),
                          lower.critraw.spatsPH=ifelse(devResPH - lower.devPH > 0,yes=1,no=0),
                          lower.critraw.spatsPhy=ifelse(devResPhy - lower.devPhy > 0,yes=1,no=0),
                          upper.critraw.spatsBio=ifelse(devResBio - upper.devBio < 0,yes=1,no=0),
                          upper.critraw.spatsPH=ifelse(devResPH - upper.devPH < 0,yes=1,no=0),
                          upper.critraw.spatsPhy=ifelse(devResPhy - upper.devPhy < 0,yes=1,no=0),
                          Q1.devBio=quantile(devResBio,probs=0.25,na.rm=TRUE) - 
                                    1.5*(quantile(devResBio,probs=0.75,na.rm=TRUE) - 
                                         quantile(devResBio,probs=0.25,na.rm=TRUE)),
                          Q1.devPH=quantile(devResPH,probs=0.25,na.rm=TRUE) - 
                                    1.5*(quantile(devResPH,probs=0.75,na.rm=TRUE) - 
                                         quantile(devResPH,probs=0.25,na.rm=TRUE)),
                          Q1.devPhy=quantile(devResPhy,probs=0.25,na.rm=TRUE) - 
                                    1.5*(quantile(devResPhy,probs=0.75,na.rm=TRUE) - 
                                         quantile(devResPhy,probs=0.25,na.rm=TRUE)),
                          Q3.devBio=quantile(devResBio,probs=0.75,na.rm=TRUE) + 
                                    1.5*(quantile(devResBio,probs=0.75,na.rm=TRUE) - 
                                         quantile(devResBio,probs=0.25,na.rm=TRUE)),
                          Q3.devPH=quantile(devResPH,probs=0.75,na.rm=TRUE) + 
                                    1.5*(quantile(devResPH,probs=0.75,na.rm=TRUE) - 
                                         quantile(devResPH,probs=0.25,na.rm=TRUE)),
                          Q3.devPhy=quantile(devResPhy,probs=0.75,na.rm=TRUE) + 
                                    1.5*(quantile(devResPhy,probs=0.75,na.rm=TRUE) - 
                                         quantile(devResPhy,probs=0.25,na.rm=TRUE)),
                          lower.critci.spatsBio=ifelse(devResBio - Q1.devBio > 0,yes=1,no=0),
                          lower.critci.spatsPH=ifelse(devResPH - Q1.devPH > 0,yes=1,no=0),
                          lower.critci.spatsPhy=ifelse(devResPhy - Q1.devPhy > 0,yes=1,no=0),
                          upper.critci.spatsBio=ifelse(devResBio - Q3.devBio < 0,yes=1,no=0),
                          upper.critci.spatsPH=ifelse(devResPH - Q3.devPH < 0,yes=1,no=0),
                          upper.critci.spatsPhy=ifelse(devResPhy - Q3.devPhy < 0,yes=1,no=0)
                    )
    
  # small plants
  # yes=1 == OK
  # no =0 == problem
  myglobal<-mutate(myglobal,
                  flagLowerRawSpats=ifelse(lower.critraw.spatsBio + lower.critraw.spatsPhy==0,yes=0,no=1),
                  flagLowerCiSpats=ifelse(lower.critci.spatsBio + lower.critci.spatsPhy==0,yes=0,no=1)
                  )
  
  # big plants
  # yes=1 == OK
  # no =0 == problem
  myglobal<-mutate(myglobal,
                  flagUpperRawSpats=ifelse(upper.critraw.spatsBio + upper.critraw.spatsPH==0,yes=0,no=1),
                  flagUpperCiSpats=ifelse(upper.critci.spatsBio + upper.critci.spatsPH==0,yes=0,no=1)
                  )
  
```

The user can save the residuals and detected outliers in an output file.

```{r export,echo=TRUE,eval=FALSE,message=FALSE, warning=FALSE}
  write.table(myglobal,"OutputFile_detectOutlierCurves_SpATS.csv",row.names=FALSE,sep="\t")
```

Summary of the cleaning procedure:

* Number of __too small__ plants according to __raw criteria__: `r ifelse(length(table(myglobal[,"flagLowerRawSpats"]))==1,
                                                                        0,table(myglobal[,"flagLowerRawSpats"])[[1]])`
* Number of __too big__ plants according to __raw criteria__: `r ifelse(length(table(myglobal[,"flagUpperRawSpats"]))==1,
                                                                        0,table(myglobal[,"flagUpperRawSpats"])[[1]])`
* Number of __too small__ plants according to __quantile criteria__: `r ifelse(length(table(myglobal[,"flagLowerCiSpats"]))==1,
                                                                        0,table(myglobal[,"flagLowerCiSpats"])[[1]]) `
* Number of __too big__ plants according to __quantile criteria__: `r ifelse(length(table(myglobal[,"flagUpperCiSpats"]))==1,
                                                                        0,table(myglobal[,"flagUpperCiSpats"])[[1]])`

## raw criteria 
```{r listing1,echo=TRUE,message=FALSE, warning=FALSE}
  cat("Lower raw outlier:")
  myglobal %>% select(Ref,genotypeAlias,repetition,Biomass24,PH24,Phy,flagLowerRawSpats) %>%
                filter(flagLowerRawSpats !=1)
   
  cat("Upper raw outlier:")
  myglobal %>% select(Ref,genotypeAlias,repetition,Biomass24,PH24,Phy,flagUpperRawSpats) %>%
                filter(flagUpperRawSpats !=1)
```

## quantile criteria 
```{r listing2,echo=TRUE,message=FALSE, warning=FALSE}
  cat("Lower quantile outlier:")
  myglobal %>% select(Ref,genotypeAlias,repetition,Biomass24,PH24,Phy,flagLowerCiSpats) %>%
                filter(flagLowerCiSpats !=1)
   
  cat("Upper quantile outlier:")
  myglobal %>% select(Ref,genotypeAlias,repetition,Biomass24,PH24,Phy,flagUpperCiSpats) %>%
                filter(flagUpperCiSpats !=1)
    
```

## raw criteria : Plots

The user can check the outlier curves detected by this procedure. We use *plant1* the dataset containing all the phenotypes of the same toy experiment as *plant4* but over thermal time. 

```{r dm,echo=TRUE,message=FALSE, warning=FALSE}
  outlierId<-filter(myglobal,flagLowerRawSpats !=1 | flagUpperRawSpats !=1)  %>% 
             select(genotypeAlias,scenario,repetition,flagLowerRawSpats,flagUpperRawSpats) %>%
             unite("geno_sce",genotypeAlias,scenario,sep="_",remove=FALSE)

  mycurve<-plant1 %>%
           unite("geno_sce",genotypeAlias,scenario,sep="_",remove=FALSE) %>%
           left_join(outlierId,mycurve,by=c("geno_sce","repetition"))
  
  # flagLower ou upper == 0 problem
  # flagLower ou upper == 1 OK
  # flag == "lower" problem en lower
  # flag == "upper" problem en upper
  # flag == "OK" plant OK
  tmp<-filter(mycurve,geno_sce %in% outlierId[,"geno_sce"]) %>%
       mutate(flagLowerRawSpats=ifelse(is.na(flagLowerRawSpats),1,flagLowerRawSpats),
         flagUpperRawSpats=ifelse(is.na(flagUpperRawSpats),1,flagUpperRawSpats),
         flag=case_when(flagLowerRawSpats == 0 ~ "lower",
                        flagUpperRawSpats == 0 ~ "upper",
                        flagUpperRawSpats + flagLowerRawSpats ==2 ~ "OK"
         ))
```
  
```{r plotoutlier,echo=TRUE,message=FALSE, warning=FALSE,fig.height=10,fig.width=10}
  ggplot(data=tmp,aes(x=thermalTime,y=biovolume,color=as.factor(repetition))) + geom_line() +
    geom_point(aes(shape=flag,color=as.factor(repetition))) + scale_shape_manual(values = c(0,19,2)) +
    facet_wrap(~geno_sce) 
```

```{r plotoutlier2,echo=TRUE,message=FALSE, warning=FALSE,fig.height=10,fig.width=10}
  ggplot(data=tmp,aes(x=thermalTime,y=plantHeight,color=as.factor(repetition))) + geom_line() +
    geom_point(aes(shape=flag,color=as.factor(repetition))) + scale_shape_manual(values = c(0,19,2)) +
    facet_wrap(~geno_sce) 
```

# Session info
```{r session,echo=FALSE,message=FALSE, warning=FALSE}
  sessionInfo()
```

# References
1. R Development Core Team (2015). R: A language and environment for statistical computing. R Foundation for
      Statistical Computing, Vienna, Austria. ISBN 3-900051-07-0, URL http://www.R-project.org.
2. Maria Xose Rodriguez-Alvarez, Martin P. Boer, Fred A. van Eeuwijk, Paul H.C. Eilers (2017). Correcting for spatial heterogeneity
  in plant breeding experiments with P-splines. Spatial Statistics URL https://doi.org/10.1016/j.spasta.2017.10.003
3. Alvarez Prado, S., Sanchez, I., Cabrera Bosquet, L., Grau, A., Welcker, C., Tardieu, F., Hilgert, N. (2019). To clean or not to clean phenotypic datasets for outlier plants in genetic analyses?. Journal of Experimental Botany, 70 (15), 3693-3698. , DOI : 10.1093/jxb/erz191 https://prodinra.inra.fr/record/481355
