---
title: "Phenoarch platform - Cleaning procedure - Point level using locfit"
author: "I.Sanchez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Phenoarch platform - Cleaning procedure - Point level using locfit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!---
Program  : detectOutlierPoints.Rmd
Author   : I.Sanchez (MISTEA)
Objective: a vignette for phisStatR library
           require phisStatR library and flagPointLocfit() function
           The user has to adapt to his own analysis.
Creation : 13/04/2018
Update   : 14/11/2019
-->

***
# Objective 

Detection of outlier points in time courses of an experiment in PhenoArch greenhouse. Use locfit smoothing function from the *locfit* library [2]. For each time course of a dataset, a locfit smoothin is applied, predictive confidence interval calculated (Y$\_$hat +/- threshold*Y$\_$hat$\_$se). Points are declared outlier if outside this confidence interval. the user choose the **threshold**.

<div style="background-color:rgba(0, 255, 0,0.1); text-align:left; vertical-align: center; padding:10px 0;">
flagPointLocfit: detection of outlier points in time courses

* @param datain input dataframe of parameters
* @param trait character, parameter of interest (ex: plantHeight)
* @param xvar character, time variable (ex: thermalTime)
* @param loopID character, ID on which to make the loop
* @param locfit.h numeric, the constant component of the smoothing parameter
* @param threshold numeric, threshold to detect on the prediction interval

@return a list:

* 1 prediction and detection of outlier on observed data
* 2 prediction on regular abscissa data
* 3 time courses with not enough point to be evaluated
</div>

```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  library(lubridate)
  library(dplyr)
  library(locfit)
  library(phisStatR)
```

# Import of data
In this vignette, we use a toy data set of the *phisStatR* library (anonymized real data set). 
```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  mydata<-plant1
  str(mydata)

  mydata<-filter(mydata,!is.na(mydata$thermalTime))
```

# Outlier points detection
I have chosen a smoothing parameter of 30 and a threshold of 8 to detect the outlier points.
```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  resu1<-flagPointLocfit(datain=mydata,trait="biovolume",xvar="thermalTime",loopID="Ref",
                         locfit.h=30,threshold=8)
```

The output report can be over-sized (more than 1Mb), for size of sub-directories in packages purposes, I choose to represent only the first genotypes...

```{r , echo = TRUE,message=FALSE, warning=FALSE,fig.height=10,fig.width=12}
  myindex<-as.character(unique(resu1[[1]][,"Ref"]))
  myindex<-myindex[1:30]
  for (i in seq(1,length(myindex),by=15)){ 
      myvec<-myindex[seq(i,i+14,1)]
      plotFlagPoint(smoothin=resu1[[1]],loopID="Ref",myselect=myvec)
  }
```

```{r , echo = TRUE,message=FALSE, warning=FALSE,fig.height=10,fig.width=12}
  filter(resu1[[1]],outlier==1)
```

# Times courses with not enough points to be evaluated
```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  # Please change the Ref column by the one in your dataframe
  if(is.null(resu1[[3]])){
    print("All the time courses have more than 4 points.")
  } else {
    ggplot(data=resu1[[3]],aes(x=x,y=y)) +
    geom_point() + facet_wrap(~Ref)
  }
```

# Session info
```{r session,echo=FALSE,message=FALSE, warning=FALSE}
  sessionInfo()
```

# References
1. R Development Core Team (2015). R: A language and environment for statistical computing. R Foundation for
      Statistical Computing, Vienna, Austria. ISBN 3-900051-07-0, URL http://www.R-project.org.
2. Catherine Loader (2013). locfit: Local Regression, Likelihood and Density Estimation.. R package version 1.5-9.1.
  https://CRAN.R-project.org/package=locfit
 

