<!---
Program  : gssAnalysisReport.Rmd
Author   : I.Sanchez (MISTEA)
Objective: a vignette for phisStatR library
           Use of smoothing splines anova to detect outlier curve on each genotype-scenario
           of an experiment with the Kullback-Leibler projection
           require phisStatR library and fitgss() function
           This program produces a pdf report and some export files (CSV format) is necessary. 
           The user has to adapt to his own analysis.
Creation : 09/12/2016
Update   : 18/10/2019
-->
---
title: "Phenoarch platform - Cleaning procedure - Curve level - gss package"
author: "I.Sanchez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Phenoarch platform - Cleaning procedure - Curve level - gss package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

***
# Objective of cleaning procedure using smoothing splines anova
Smoothing spline analysis of variance on each genotype-scenario of an experiment. Detection of outlier repetition if significant TT*Rep (thermal time by repetition) interaction using a Kullblack-Leibler projection (KL). I consider a genotype-scenario as outlier:

* biovolume: if KL > 0.05
* plantHeight: if KL > 0.05
* leafArea: if KL > 0.05

The input dataset must contain the following columns:

* experimentAlias
* genotypeAlias
* scenario
* repetition
* thermalTime (for thermal time)
* parameter of interest (biovolume, plantHeight etc...)

The five first column names are standard names extracted from the web service.

# Import of data
```{r,echo=TRUE,message=FALSE, warning=FALSE}
  library(ggplot2)
  library(lubridate)
  library(tidyr)
  library(dplyr)
  library(gss)
  library(phisStatR)

  myreport<-substr(now(),1,10)
```

```{r,echo=TRUE,message=FALSE, warning=FALSE}
  # Import data, here is a dataset in the phisStatR package, You have to import your own dataset
  # using a read.table() statement or a request to the web service
  # You can add some datamanagement statements...
  #------------------------------------------------------------------------
  # Please, add the 'Ref' and 'Genosce' columns if don't exist. 
  # 'Ref' is the concatenation of experimentAlias-Line-Position-scenario
  # 'Genosce' is the concatenation of experimentAlias-genotypeAlias-scenario
  #------------------------------------------------------------------------
  
  mydata<-unite(plant3,Genosce,experimentAlias,genotypeAlias,scenario,sep="-",remove=FALSE)
  mydata<-arrange(mydata,Genosce)
  
```

```{r oneprint,echo=FALSE,message=FALSE, warning=FALSE}
  cat("-------------- plant3 dataset ---------------\n")  
  printExperiment(datain=mydata,manip="manip3")
```

```{r algo,echo=TRUE,message=FALSE, warning=FALSE}
  # For one parameter, for example biovolume
  resbio<-fitGSS(datain=mydata,trait="biovolume",loopId="Genosce") 
```

# Curves by genotype-scenario

## Biovolume
```{r,echo=TRUE,message=FALSE, warning=FALSE}
  outlierbio<-printGSS(object=resbio[[2]],threshold = 0.05)
  klbio<-printGSS(object=resbio[[2]],threshold = NULL)
  
  cat("Detection of outlier curve with KL projection:\n")
  print(outlierbio)
  
  #------------------------------------------------
  # You can export these two datasets
  # suppress the comments
  #------------------------------------------------
  #write.table(outlierbio,paste0(myreport,"outlier_gss_biovolume.csv"),row.names = FALSE,sep="\t")
  #write.table(klbio,paste0(myreport,"KLprojection_gss_biovolume.csv"),row.names = FALSE,sep="\t")
```

I take a threshold of 0.05 for this example. We can take a more conservative threshold like 0.01 or 0.02 to detect more outlier curves...
    
```{r graph,echo=TRUE,message=FALSE, warning=FALSE,fig.width=12, fig.height=16}
  # plot of the smoothing splines by genotype-scenario  
  for(i in seq(1,length(unique(mydata[,"Genosce"])),by=12)){ 
    myvec<-seq(i,i+11,1)
    myvec<-myvec[myvec<=length(unique(mydata[,"Genosce"]))]
    print(plotGSS(datain=mydata,modelin=resbio[[1]],trait="biovolume",myvec=myvec,lgrid=50))
    cat("\n\n")
  }

```

# Session info

```{r session,echo=FALSE,message=FALSE, warning=FALSE}
  sessionInfo()
```

# References
1. R Development Core Team (2015). R: A language and environment for statistical computing. R Foundation for
      Statistical Computing, Vienna, Austria. ISBN 3-900051-07-0, URL http://www.R-project.org.
2. Chong Gu (2014). Smoothing Spline ANOVA Models: R Package gss. Journal of Statistical Software, 58(5), 1-25. URL
  http://www.jstatsoft.org/v58/i05/.


